Chatbots and messaging platforms in the classroom: an analysis from the teacher's perspective

### 2021-11-08 edubots_surveys.Rnw
Some comments deleted, also reflow #50

A \textit{chatbot} is a program, sometimes developed using Artificial Intelligence techniques, able to communicate in a similar way as humans do \cite{Gong2008}, using, in some cases, natural language. Indeed, most of these applications would be, apparently, close to passing a classic Turing test \cite{moor2003turing}, since they are able to answer almost any question fluently, and even pose own questions in their conversations. The recent advancement in this discipline has played an important role in many fields especially in education and online tutoring \cite{clarizia2018,Smutny2020}. These automatic systems facilitate the delivery of personalized learning by adapting to students’ pace of learning and providing customized online tutoring outside the classroom. Thus, chatbots can significantly contribute to providing interactive learning experiences as well as improving individual attention \cite{agarwal2020}.
 In this sense, chatbot technology offers a great opportunity for the improvement of tutoring systems \cite{Daniel2016,agarwal2020}, as not all students are comfortable with face-to-face tutoring with the instructor. In fact, in many cases, some students experience stress because of their need to ask a question in front of the entire class. This is why many students resort to silence  
 % Antonio - I guess refrain to silence is the opposite of what we wanted to say, isn't it? :D
 % mistaking "resort" for "refrain". You don't say "refrain to", but refrain from.
 in the classroom and tend to contact their teachers later via email.
 Consequently, this can lead to not only a delay in obtaining answers to their questions but can also cause a significant increase in their teachers’ workload, especially when factoring in the disproportion between the class size and the number of teachers. Hence, the chatbot technology has a potential to mitigate this problem by providing answers to students’ questions and facilitating a dynamic and autonomous learning experience \cite{Griol2014,kim2020}. In addition, using an automated system such as a chatbot, can draw teachers’ attention to topics that students struggle to comprehend or need further assistance in understanding.

@@

Our intention from the beginning was to try to gather a sufficient amount of responses in all sectors, as well as cover different sectors of higher education. In doing so, an online survey was administered (i.e., Google Form), in Spanish, which was intended to be answered in a very short amount of time. Initially, our intention was to mirror the questions used in \cite{MoraChatbots2021:anon}; 
% Antonio - I would say something like: "but from teachers' perspective"
however, an initial exploration of results led us to create a second survey. The survey items focused on the use of messaging apps in the teaching practice, the type of chatbot use cases that teachers consider useful to be used in their teaching, and the type of interactions that teachers prefer to have with their students while using messaging apps in their teaching.

% Move a part of this to an appendix: Noorhan
The {\bf initial survey} was done in two separate Google forms, one for university and other for tertiary non-university teachers. Questions and responses were the same, except for the type of tertiary degrees that were taught. In both cases, we used mainly mailing lists, as well as Telegram groups, to spread the link to the form. The university form was spread in Spanish (mainly Andaluc\'{\i}a and Galicia) universities, and also Costa Rica and Mexico. The tertiary non-university teachers are based mainly in Andaluc\'{\i}a. Dissemination and answers took place in the first quarter of 2021, post-pandemic and while, at least in Spain, many universities had mandatory virtual teaching.

@@

Overall, the majority of the teachers responded that they do use messaging apps in their classrooms, from which apps provided by the academic institution (n=159, 56\%) and WhatsApp (n=124, 44.0\%) were the most common responses. Only 19 teachers (6.7\%) replied that they do not use any messaging app in their class. %Remember to add something to the discussion below - JJ
%
\begin{table}[h!tbp]
\begin{tabular}{lllll}
\hline
\multicolumn{1}{c}{Messaging App} & \multicolumn{2}{c}{Yes} & \multicolumn{2}{l}{No} \\ \cline{2-5} 
\multicolumn{1}{c}{} & \multicolumn{1}{c}{Frequency} & \multicolumn{1}{c}{\%} & Frequency & \multicolumn{1}{c}{\%} \\ \hline
Telegram & 62 & 22.0 & 220 & 78.0 \\
WhatsApp & 124 & 44.0 & 158 & 56.0 \\
Slack & 15 & 5.3 & 267 & 94.7 \\
Other & 60 & 21.3 & 222 & 78.7 \\
\begin{tabular}[c]{@{}l@{}}Provided by the Academic\\ Institution\end{tabular} & 159 & 56.4 & 123 & 43.6 \\
None & 19 & 6.7 & 263 & 93.3 \\ \hline
\end{tabular}
\vspace{0.8em}
\caption{Use of messaging apps to assist the learning process.}
\label{tab:use_messapps}
\end{table}

@@

Although no significant differences were found regarding gender, female teachers answered they use instant messaging apps more than male teachers (about 10\% more). Also, teachers in vocational Education use WhatsApp more than university teachers.
Regarding the distribution of the use of messaging apps per age, there are no significant differences for WhatsApp and apps provided by their own academic institutions. % Olia requests support for this.
However, younger teachers also use Telegram with more than 25\%, a percentage that falls to about 10\% for teachers that are 55 or older. %The third column compares the use of these apps considering the number of years of experience. 
One interesting result is that about 65\% of teachers with more than 25 years of experience use the platforms provided by their institutions while the percentage goes down to less than 50\% for teachers with 6-15 years of experience.

@@

Some of the questions in the survey were focused on the impact of the COVID19 pandemic between 2020 and 2021, in the teachers' viewpoints towards the use of instant messaging apps in their class. Our main intention in this case was to assess whether a crisis will bring about some kind of change in the use of tools.
The 282 
% Fran, please check Olia's comment as having 282 answers 
% -> I did remove some responses that Olia considered 'No' by default (i guess). I' ll keep it to 282 to be consistent
answers are summarized in Figs. \ref{figure:gender*chatbots_use_postcovid}, \ref{figure:sector*chatbots_use_postcovid}, \ref{figure:years*chatbots_use_postcovid}, \ref{figure:disc*chatbots_use_postcovid}, showing that about 77\% of teachers already used these tools before the pandemic and kept using them during the pandemic lockdowns that forced students and educators to use remote education schemes. Moreover, approximately 15\% of them switched their messaging app for one that offered a safer interaction with their students. According to the responses, an additional 16\% started using messaging apps during the pandemic for the first time in their classes. Aggregating the answers, currently 91\% of the teachers are using messaging apps as learning tools. 

@@

Finally, statistical significance was found for the discipline and post-covid changes. Regarding disciplines (see Fig. \ref{figure:discipline*postcovid_chisq}), the number of educators from Humanities and Social Sciences that started using messaging apps after the pandemic is larger than expected and these number of responses explains about 19\% of the chi-square test results. Also, the number of educators from Technology or Engineering that started using safer alternatives are less than expected (compared to the educators from other disciplines). Bear in mind that some disciplines were merged to avoid very low number of responses for some of the cells.

@@

From a qualitative point of view, these results show consensus in a few key aspects:
\begin{itemize}
 \item Teachers want to have support from their corresponding institution. This is probably due to the strict European data regulations and to avoid the overhead required to sing up and deleting students from year to year.

 \item Teachers need sustainability. The main reason is that implementing changes is expensive in terms of effort and prone to errors in the first years of implementing new procedures, therefore, it is logical that if chatbots or messaging platforms are to be introduced, they want this change to be as permanent as possible.

 \item Keep it Simple. Media interaction requirements are simple and only a 2.4\% is requesting more features. This could be a sign of technological burn out produced by the previous and current courses where, due to the pandemic, the use of computer has increased significantly. 
\end{itemize}

>>>>

A \textit{chatbot} is a program, sometimes developed using Artificial
Intelligence techniques, able to communicate in a similar way as humans do
\cite{Gong2008}, using, in some cases, natural language. Indeed, most of these
applications would be, apparently, close to passing a classic Turing test
\cite{moor2003turing}, since they are able to answer almost any question
fluently, and even pose own questions in their conversations. The recent
advancement in this discipline has played an important role in many fields
especially in education and online tutoring
\cite{clarizia2018,Smutny2020}. These automatic systems facilitate the delivery
of personalized learning by adapting to students’ pace of learning and providing
customized online tutoring outside the classroom. Thus, chatbots can
significantly contribute to providing interactive learning experiences as well
as improving individual attention \cite{agarwal2020}.  In this sense, chatbot
technology offers a great opportunity for the improvement of tutoring systems
\cite{Daniel2016,agarwal2020}, as not all students are comfortable with
face-to-face tutoring with the instructor. In fact, in many cases, some students
experience stress because of their need to ask a question in front of the entire
class. This is why many students resort to silence in the classroom and tend to
contact their teachers later via email.  Consequently, this can lead to not only
a delay in obtaining answers to their questions but can also cause a significant
increase in their teachers’ workload, especially when factoring in the
disproportion between the class size and the number of teachers. Hence, the
chatbot technology has a potential to mitigate this problem by providing answers
to students’ questions and facilitating a dynamic and autonomous learning
experience \cite{Griol2014,kim2020}. In addition, using an automated system such
as a chatbot, can draw teachers’ attention to topics that students struggle to
comprehend or need further assistance in understanding.

@@

Our intention from the beginning was to try to gather a sufficient amount of
responses in all sectors, as well as cover different sectors of higher
education. In doing so, an online survey was administered (i.e., Google Form),
in Spanish, which was intended to be answered in a very short amount of
time. Initially, our intention was to mirror the questions used in
\cite{MoraChatbots2021:anon}; however, an initial exploration of results led us
to create a second survey. The survey items focused on the use of messaging apps
in the teaching practice, the type of chatbot use cases that teachers consider
useful to be used in their teaching, and the type of interactions that teachers
prefer to have with their students while using messaging apps in their teaching.

The {\bf initial survey} was done in two separate Google forms, one for
university and other for tertiary non-university teachers. Questions and
responses were the same, except for the type of tertiary degrees that were
taught. In both cases, we used mainly mailing lists, as well as Telegram groups,
to spread the link to the form. The university form was spread in Spanish
(mainly Andaluc\'{\i}a and Galicia) universities, and also Costa Rica and
Mexico. The tertiary non-university teachers are based mainly in
Andaluc\'{\i}a. Dissemination and answers took place in the first quarter of
2021, post-pandemic and while, at least in Spain, many universities had
mandatory virtual teaching.

@@

Overall, the majority of the teachers responded that they do use messaging apps
in their classrooms, from which apps provided by the academic institution
(n=159, 56\%) and WhatsApp (n=124, 44.0\%) were the most common responses. Only
19 teachers (6.7\%) replied that they do not use any messaging app in their
class.
%
\begin{table}[h!tbp]
\begin{tabular}{lllll}
\hline
\multicolumn{1}{c}{Messaging App} & \multicolumn{2}{c}{Yes} & \multicolumn{2}{l}{No} \\ \cline{2-5} 
\multicolumn{1}{c}{} & \multicolumn{1}{c}{Frequency} & \multicolumn{1}{c}{\%} & Frequency & \multicolumn{1}{c}{\%} \\ \hline
Telegram & 62 & 22.0 & 220 & 78.0 \\
WhatsApp & 124 & 44.0 & 158 & 56.0 \\
Slack & 15 & 5.3 & 267 & 94.7 \\
Other & 60 & 21.3 & 222 & 78.7 \\
\begin{tabular}[c]{@{}l@{}}Provided by the Academic\\ Institution\end{tabular} & 159 & 56.4 & 123 & 43.6 \\
None & 19 & 6.7 & 263 & 93.3 \\ \hline
\end{tabular}
\vspace{0.8em}
\caption{Use of messaging apps to assist the learning process.}
\label{tab:use_messapps}
\end{table}

@@

Although no significant differences were found regarding gender, female teachers
answered they use instant messaging apps more than male teachers (about 10\%
more). Also, teachers in vocational Education use WhatsApp more than university
teachers. Regarding the distribution of the use of messaging apps per age, there
are no significant differences for WhatsApp and apps provided by their own
academic institutions. % Olia requests support for this. 
However, younger teachers also use Telegram with more than 25\%, a percentage
that falls to about 10\% for teachers that are 55 or older. %The third column
                                %compares the use of these apps considering the
                                %number of years of experience.
One interesting result is that about 65\% of teachers with more than 25 years of
experience use the platforms provided by their institutions while the percentage
goes down to less than 50\% for teachers with 6-15 years of experience.

@@

Some of the questions in the survey were focused on the impact of the COVID19
pandemic between 2020 and 2021, in the teachers' viewpoints towards the use of
instant messaging apps in their class. Our main intention in this case was to
assess whether a crisis will bring about some kind of change in the use of
tools. The 282 answers are summarized in
Figs. \ref{figure:gender*chatbots_use_postcovid},
\ref{figure:sector*chatbots_use_postcovid},
\ref{figure:years*chatbots_use_postcovid},
\ref{figure:disc*chatbots_use_postcovid}, showing that about 77\% of teachers
already used these tools before the pandemic and kept using them during the
pandemic lockdowns that forced students and educators to use remote education
schemes. Moreover, approximately 15\% of them switched their messaging app for
one that offered a safer interaction with their students. According to the
responses, an additional 16\% started using messaging apps during the pandemic
for the first time in their classes. Aggregating the answers, currently 91\% of
the teachers are using messaging apps as learning tools.

@@

Finally, statistical significance was found for the discipline and post-covid
changes. Regarding disciplines (see
Fig. \ref{figure:discipline*postcovid_chisq}), the number of educators from
Humanities and Social Sciences that started using messaging apps after the
pandemic is larger than expected and these number of responses explains about
19\% of the chi-square test results. Also, the number of educators from
Technology or Engineering that started using safer alternatives are less than
expected (compared to the educators from other disciplines). Bear in mind that
some disciplines were merged to avoid very low number of responses for some of
the cells.

@@

From a qualitative point of view, these results show consensus in a few key aspects:
\begin{itemize}
 \item Teachers want to have support from their corresponding institution. This
   is probably due to the strict European data regulations and to avoid the
   overhead required to sing up and deleting students from year to year.
 \item Teachers need sustainability. The main reason is that implementing
   changes is expensive in terms of effort and prone to errors in the first
   years of implementing new procedures, therefore, it is logical that if
   chatbots or messaging platforms are to be introduced, they want this change
   to be as permanent as possible.
 \item Keep it Simple. Media interaction requirements are simple and only a 2.4\% is requesting more features. This could be a sign of technological burn out produced by the previous and current courses where, due to the pandemic, the use of computer has increased significantly.
\end{itemize}
